What's in a name? Disambiguation strategies for relational descriptions

### 2023-03-19 2023 Disambiguation/Disambiguation.tex
work in progress on Disambiguation article

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx,stmaryrd}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{bbold,bbm}
\makeatletter
\def\amsbb{\use@mathgroup \M@U \symAMSb}
\makeatother
\usepackage[cal=boondoxo]{mathalfa}
\newcommand{\typesemi}{\mathbin{\text{\textbb{;}}}}
\newcommand{\typesubset}{\mathbin{\mathbbm{\sqsubseteq}}}
\newcommand{\typeunion}{\mathbin{\mathbb{U}}}
\newcommand{\typeinter}{\mathbin{\text{\rotatebox[origin=c]{180}{$\typeunion$}}}}
\newcommand{\typevee}{\mathbbm{T}}
\newcommand{\typenil}{\mathbbm{F}}
\newcommand{\typecomp}[1]{\neg{#1}}
\newcommand{\typeconv}[1]{{#1}^\smallsmile}
\newcommand{\typeident}{\mathbb{1}}
\newcommand{\typetyped}[2]{{#1}_{\llbracket #2 \rrbracket}}
\newcommand{\conv}[1]{{#1}^{-1}}
\usepackage{algorithmic}
\begin{document}

@@

The type of a term in $\mathcal{R}$ is given by a partial typing function, which we define inductively via the relation symbol $:$ as follows:
\begin{align*}
x_{A*B} &: A * B & \text{For $x \in \mathcal{N} \cup \{\epsilon, T, F\}$ and $A,B\in \mathcal{T}$}\\
1_{A} &: A * A&\\
\varphi \odot \psi &: A * B &\quad\text{For $\odot \in \{\cup,\cap\}$, $\varphi : A * B$ and $\psi : A * B$}\\
\varphi ; \psi &: A * B &\text{For $\varphi : A * C$ and $\psi : C * B$}\\
\overline{\psi} &: A * B &\text{For $\psi : A * B$}\\
\conv{\psi} &: A * B &\text{For $\psi : B * A$}
\end{align*}
We say that a term $r\in \mathcal{R}$ is \emph{well-typed} if $r : A*B$ for some $A$ and $B$.
By induction on the inductive definition of $\mathcal{R}$, the relation $:$ is between $\mathcal{R}$ and types $\mathcal{T}\times\mathcal{T}$ and univalent.
In other words: well-typed terms in $\mathcal{R}$ have precisely one type, all of its sub-terms are well-typed, and $\bot$ does not occur as a pseudo-concept.

@@

\paragraph{[Aside],} The Ampersand language also allows for \verb=CLASSIFY A IS B /\ C= statements, meaning that elements are in $A$ if and only if they are in $B$ and in $C$.
The information about this captured in $\preceq$ is just that $A \preceq B$ and $A\preceq C$, which means that information is lost.
We can consider the lost information when dealing with the $\typesemi$ operation by changing the $\epsilon$ relations introduced there.
However, it is an open problem to deal with this with the $\typeinter$ operation, so we give a simpler presentation that ignores this additional information in \verb=CLASSIFY _ IS _= statements for now.

>>>>

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx,stmaryrd}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{bbold,bbm}
\makeatletter
\def\amsbb{\use@mathgroup \M@U \symAMSb}
\makeatother
\usepackage[cal=boondoxo]{mathalfa}

\DeclareMathAlphabet{\pazocal}{OMS}{zplm}{m}{n}
\newcommand{\powerset}{\pazocal{P}}

\newcommand{\typesemi}{\mathbin{\text{\textbb{;}}}}
\newcommand{\typesubset}{\mathbin{\mathbbm{\sqsubseteq}}}
\newcommand{\typeunion}{\mathbin{\mathbb{U}}}
\newcommand{\typeinter}{\mathbin{\text{\rotatebox[origin=c]{180}{$\typeunion$}}}}
\newcommand{\typevee}{\mathbbm{T}}
\newcommand{\typenil}{\mathbbm{F}}
\newcommand{\typecomp}[1]{\neg{#1}}
\newcommand{\typeconv}[1]{{#1}^\smallsmile}
\newcommand{\typeident}{\mathbb{1}}
\newcommand{\typetyped}[2]{{#1}_{\llbracket #2 \rrbracket}}
\newcommand{\conv}[1]{{#1}^{-1}}
\usepackage{algorithmic}
\begin{document}

@@

The type of a term in $\mathcal{R}$ is given by a partial typing function, which we define inductively via the relation symbol $:$ as follows:
\begin{align*}
x_{A*B} &: A * B & \text{For $x \in \mathcal{N} \cup \{\epsilon, T, F\}$}\\
1_{A} &: A * A&\\
\varphi \odot \psi &: A * B &\quad\text{For $\odot \in \{\cup,\cap\}$, $\varphi : A * B$ and $\psi : A * B$}\\
\varphi ; \psi &: A * B &\text{For $\varphi : A * C$ and $\psi : C * B$}\\
\overline{\psi} &: A * B &\text{For $\psi : A * B$}\\
\conv{\psi} &: A * B &\text{For $\psi : B * A$}
\end{align*}
We say that a term $r\in \mathcal{R}$ is \emph{well-typed} if $r : A*B$ for $A,B\in \mathcal{T}$ and every sub-term is well-typed.
By induction on the inductive definition of $\mathcal{R}$, the relation $:$ is univalent.
In other words: well-typed terms in $\mathcal{R}$ have precisely one type, all of its sub-terms are well-typed, and $\bot$ does not occur as a pseudo-concept.

@@

\paragraph{[Aside]} The Ampersand language also allows for \verb=CLASSIFY A IS B /\ C= statements, meaning that elements are in $A$ if and only if they are in $B$ and in $C$.
The information about this captured in $\preceq$ is just that $A \preceq B$ and $A\preceq C$, which means that information is lost.
We can consider the lost information when dealing with the $\typesemi$ operation by changing the $\epsilon$ relations introduced there.
However, it is an open problem to deal with this with the $\typeinter$ operation, so we give a simpler presentation that ignores this additional information in \verb=CLASSIFY _ IS _= statements for now.

Translating a term from $\mathcal{D}$ to $\mathcal{R}$ proceeds by considering which choices are type correct.
However, if we were to simply implement the simple algorithm proposed earlier, too many terms would be deemed ambiguous:
the term $\typetyped{\typeident}{\mathrm{Event}} \typesemi \typevee \typesemi \typetyped{\mathrm{scheduled}}{\mathrm{Event}*\mathrm{Time}}$ with $\mathrm{Task}\preceq\mathrm{Event}$ would not disambiguate to the obvious term, since $\typetyped{\typeident}{\mathrm{Event}} \typesemi \typetyped{\typevee}{\mathrm{Task}*\mathrm{Task}} \typesemi \typetyped{\mathrm{scheduled}}{\mathrm{Event}*\mathrm{Time}}$ is also well-typed.
To deal with this, we prefer terms in which the $\epsilon$ relations have the same source and target.

Given a term $D\in\mathcal{D}$, the function $p(D)$ will return a set of types and a preference relation between those types.
The preference relation is a pre-order.
We will write a tree node as $t \mapsfrom \{c_1,\ldots,c_n\}$ where the child-nodes $c_i$ are themselves written as trees, and $t$ is a set of types that the corresponding term could have.
For example, if $p(D) = t_1 \mapsfrom \{t_2 \mapsfrom \{t_3 \mapsfrom \{\}\},t_4 \mapsfrom \{\}\}$, then types of $t_1$ are preferred over those in $t_2$, $t_3$ and $t_4$, those of $t_2$ are preferred over $t_3$, and those types in $t_4$, are incomparable to those in $t_2$ and $t_3$.

We define a binary operation $\times$ on such trees as follows:
\[
(a \mapsfrom A) \times (b \mapsfrom B)
= a \cap b \mapsfrom 
\]

We define $p$ as follows:
\begin{align*}
p(\xi) &= \{{A*B} \mid \xi : (A,B)\}\mapsfrom\{\}\\
p(\Xi) &= \{{A*A} \mid \xi = \epsilon_{A*A}\}\mapsfrom \{\{{A*B} \mid \xi : {A*B} \wedge (A\neq B \vee \xi \neq \epsilon_{A*B})\} \mapsfrom\{\}\}\\
p(\typetyped{\varphi}{A*B}) &= \{A*B \mid A*B\in p(\varphi)[i] \text{ for some $i$}\}\mapsfrom\{\}\\
p(\varphi \cup \psi) &= ..?\\
p(\varphi \cap \psi) &= p(\varphi \cup \psi)\\
p(\varphi ; \psi) &= ..?\\
p(\overline{\varphi}) &= p(\varphi)\\
p(\conv{\varphi}) &= [\{B*A\mid A*B\in p(\varphi)[i]\}]_i
\end{align*}

\paragraph{[Aside]}
Without optimizations, the complexity of the algorithm for constructing these trees is exponential: the combination of two trees for $\cap$, $\cup$ and $;$ will be of a size that is the product of the two.
However, we hope to be able to reduce the size of trees by looking for equivalent trees that express the same preferences.
If trees can be reduced so that the number of nodes never exceeds the number of types in them, this should give the tree construction polynomial complexity.
